Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-25 paper/paper.tex
remove unused import
save some space

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
%% \citestyle{acmauthoryear}   %% For author/year citations
\citestyle{acmnumeric}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{xspace}
\usepackage[nounderscore]{syntax}
\usepackage{thmtools} % needed for sane cleveref naming of theorems
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}

@@

\usepackage{listings}
\lstdefinelanguage{Agda}{
keywords={data, where, module, import, open, public,
record, field, let, in, if, then, else, case, of, with,
do, postulate, primitive, mutual, abstract, private,
forall, exists, cong, set, prop, sort, Level, Data, Type,
Renamer},
morekeywords=[2]{Set, Type, Prop, moduleFuncs, instance, Named},
sensitive=true,
comment=[l]{--},
morecomment=[s]{\{-}{-\}},
morestring=[b]",
mathescape=true,
escapeinside={(*@}{@*)}
}
\lstset{
language=Agda,
basicstyle=\ttfamily\small,
keywordstyle=\color{blue},
keywordstyle=[2]\color{teal},
identifierstyle=\color{black},
commentstyle=\color{gray}\textit,
stringstyle=\color{orange},
numbers=none,
numberstyle=\tiny\color{gray},
stepnumber=1,
numbersep=5pt,
showstringspaces=false,
tabsize=4,
captionpos=b,
breaklines=true,
breakatwhitespace=false,
rulecolor=\color{black},
texcl=true,
backgroundcolor=\color{backcolour},
frame=single
}

@@

\newcommand{\Subst}{\textrm{Subst}}
\newcommand{\SPF}{SPF}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\left\llbracket{#1}\right\rrbracket}
\newcommand{\semg}[1]{\sem{#1}\gamma}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\StringSem}{\mathbf{String}}
\newcommand{\CharGram}{\texttt{Char}}
\newcommand{\StringGram}{\texttt{String}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\semcat}{\mathbf{C}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

\begin{figure}
\begin{lstlisting}
f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
f (a , b) = inl (a , b)
\end{lstlisting}\begin{mathpar}
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {~}
      {a : \literal a \vdash a : \literal a}
      \\
      \inferrule
      {~}
      {b : \literal b \vdash b : \literal b}
    }
    {a : \literal a , b : \literal b \vdash (a , b) : \literal a \otimes
      \literal b}
  }
  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a , b) :
    (\literal a \otimes \literal b) \oplus \literal c}
\end{mathpar}
\caption{(\Agda{fillerlink}) \stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
}
\label{fig:fingram}
\end{figure}
For this interpretation of parse trees as terms to make sense, our
calculus cannot allow for \emph{any} of the usual structural rules of
type theory: weakening, contraction and exchange. Weakening allows
for variables to go unused, while contraction allows for the same
variable to be used twice, but in a parse tree, every character must
be accounted for exactly once. That is, we want to prevent the following
erroneous derivations,
\[
  a : \literal a , b : \literal b \not \vdash a : \literal a \qquad a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a
\]
%
Finally, the ordering of characters in a string cannot be ignored while
parsing, so we omit the exchange rule because it would allow
for variables in the context to be reordered,
\[
  a : \literal a , b : \literal b \not \vdash (b , a) : \literal b \otimes \literal a
\]
% otherwise we would have that $\inl(x\otimes y)$ is a valid
% term in context $y:b,x:a$, but there should be no parses of $ba$ in
% this grammar.

@@

\begin{figure}
\begin{lstlisting}
data A(*@$^*$@*) : L where
  nil : (*@$\uparrow$@*)(A(*@$^*$@*))
  cons : (*@$\uparrow$@*)(A (*@$\lto$@*) A(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
\end{lstlisting}
% \begin{align*}
% \data &~A^{*} : L~\where\\
%       & \nil : \ltonl {A^{*}} \\
%       & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
% \end{align*}
\caption{Kleene Star as an inductive type}
\label{fig:kleenestarinductive}
\end{figure}

@@

At the bottom of
\cref{fig:kleeneabstractproof}, we give a defintion of the same function as it
appears in our current Agda implementation \cite{?}. $h'$ is written as a
composition of combinators,
each without any pattern matching or introduction
of named linear variables. More discussion of our implementation is given in \cref{subsec:impl}.

\begin{figure}
\footnotesize
\begin{mathpar}
  \inferrule
  {
    \inferrule
    {~}
    {\cdot \vdash \nil : \A^*}
    \\
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {~}
        {a_1 : \A, a_2 : \A , as : \A^* \vdash \cons~a_1 (\cons~a_2~as) : \A^*}
      }
      {aa : \A \otimes \A , as : \A^* \vdash \letin {(a_1 , a_2)}
        {aa} {\cons~a_1 (\cons~a_2~as)} : \A^*}
    }
    {\cdot \vdash f := \lamblto {aa} {\lamblto {as} {\letin {(a_1 , a_2)}
        {aa} {\cons~a_1 (\cons~a_2~as)}}} : (\A \otimes \A) \lto \A^* \lto \A^*}
  }
  {aas : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(aas) : \A^* }
\end{mathpar}
\begin{lstlisting}
h : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
h nil = nil
h (cons (a1 , a2) as) = cons a1 (cons a2 (h as))
\end{lstlisting}
\begin{lstlisting}
h' : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
h' = fold*r nil (cons (*@$\circ$@*) id (*@$\otimes$@*) cons (*@$\circ$@*) assoc(*@$^{-1}$@*))
\end{lstlisting}
\caption{(\Agda{\fillerlink}) A parse transformer for abstract grammars}
\label{fig:kleeneabstractproof}
\end{figure}

@@

\begin{axiom}[Distributivity]
\label{ax:dist}
For any $A : (x : X) \to Y x \to L$, the function
\[
 \lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}} : \ltonl {\left(  \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
\LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}} \right)}
\]
has an inverse.
\end{axiom}

@@

  We demonstrate the weak equivalence between $Parse_{N}$ and
  $Parse_{D}$ by constructing parse transformers between the two
  grammars. To build the parse transformer
  $\ltonl{(Parse_{N} \lto Parse_{D})}$, we strengthen our inductive hypothesis
  and build a term
\newcommand{\NtoD}{\texttt{NtoD}}
\newcommand{\DtoN}{\texttt{DtoN}}
  \[
    \NtoD : \ltonl {\left(\Trace_{\N}~\s~\true \lto \bigwith\limits_{\X : \D.\states} \bigwith\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\D}~\X~\true \right)}
  \]
  that maps a trace in $\N$ from an arbitrary state $\s$ to a trace in $\D$
  that may begin at any subset of states $\X$ that contains $\s$. $\NtoD$ may then be instantiated at
  $\s = \N.\init$ and $\X = D.\mathsf{init}$ to get the desired
  parse transformer.

@@

To construct a term from DFA traces to NFA traces, we similarly strengthen our
inductive hypothesis and build a parse transformer
\[
  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \bigoplus\limits_{s : \N.\states}\bigoplus\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\N}~s~\true \right)}
\]
Because the states of $D$ are $\epsilon$-closed subsets of $N.\states$, if two
states $s$ and $s'$ belong to the the same $\epsilon$-closed subset
$X : \mathbb{P}_{\epsilon}(N.\states)$ then there exists a $\epsilon$-path in
$N$ between the two, but we do not necessarily know which one. Similarly, the
data contained in the type $\Trace_{\D}~X~true$ is the \emph{existence} of an accepting trace in
$\N$ beginning at some state in $\D$.

To define $\DtoN$, we need a choice function that extracts out a trace in $N$
from the mere existence of one. We achieve this by choosing the smallest trace
through $\N$ subject to an ordering on the non-linear types $N.\states$,
$N.\transitions$, and $N.\epstransitions$. In essence, the given orderings
specify a global disambiguation strategy.

%% For a moment, let us turn our attention to the content behind the traces in $\D$.
%% Traces from subset $\X$ to subset $\Y$ describes the \emph{existence} of some trace
%% in the $\N$ from some $\x \in \X$ to some $\y \in \Y$. In general it is not clear how
%% to extract out a choice of trace in $\N$, which is what $\DtoN$ is asking of us.
%% However, we may further take the states and transitions of $\N$ to be \emph{ordered}.
%% This allows us to
%% induce an order on traces in $\D$ from $\X$ to $\Y$. From these orderings we can then
%% define the necessary choice functions on each type by taking the smallest with
%% respect to the ordering.

@@

%% Next if the trace in $\D$ is defined via transition
%% $\X \overset{\stringquote{c}}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
%% $\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
%% $\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
%% there exists some $\s' \in \X$ such that $\s'$ transitions to $\s$ by some number of
%% $\epsilon$-transitions followed by a transition labeled by $\c$.
%% We order all
%% such traces of this form and choose the smallest one, which is then linearly
%% combined with the parse of $\c$ to build an $\N$ trace. This concludes the
%% definition of $\DtoN$, thus showing that $Parse_{N}$ and $Parse_{D}$ are
%% weakly equivalent.
\end{proof}

Finally, given any regular expression we can construct a
\emph{strongly} equivalent NFA. While only weak equivalence is
required to construct a parser, proving the strong equivalence shows
that other aspects of formal grammar theory are also verifiable in
\theoryabbv.
\newcommand{\R}{\texttt{R}}
\begin{construction}[Thompson's Construction, \Agda{\fillerlink}]
  \label{cons:thompson}
  Given a regular expression $\R$, we build an NFA $\N$ such that $\R$ is
  strongly equivalent to $\Trace_{N}(\N.\init)$.
\end{construction}
\begin{proof}
We use a variant of Thompson's construction
\cite{thompsonProgrammingTechniquesRegular1968}, showing that NFAs are, up to strong equivalence, closed under each type operation for regular expressions.

%% In this paper we will walk through
%% some of the cases of this proof. The remaining cases
%% follow
%% similar reasoning and the details may be found in the formalized artifact.

@@

%% In this sketch, we elide the proof that these maps do indeed form homomorphisms.
%% Further, we are eliding the needed NFAs for the cases where the regular
%% expression is $I$, $0$, a Kleene star, or a concatenation. All of these
%% elided details may be found in the Agda formalization.
\end{proof}

@@

While we have shown only examples for context-free grammars, in fact
arbitrarily complex grammars are encodable in \theoryabbv. To
demonstrate this, we show that for any \emph{non-linear} function $P :
\StringSem \to U$, where here $\StringSem$ is the \emph{non-linear} type
of strings over the alphabet, we can construct a grammar whose parses
correspond to $P$.
%% Define $\StringSem$ to be the non-linear list type over the alphabet $\Sigma$ and
%% let  be a mapping from the set of strings to the universe of
%% nonlinear types. The function $P$ describes a family of nonlinear types
%% dependent on $\StringSem$.
%% We view the image of $P$ as a set of parse trees for each input string, and thus
%% $P$ carries the structure of a formal grammar. However, even though $P$ describes this grammatical data, it is not
%% a first-class linear type in \theoryabbv. 
\[
Reify~P = \LinSigTyLimit {w} {\StringSem} {\LinSigTyLimit {x} {P~w} {~\lceil w \rceil}}
\]
%% \steven{Probably put this ceiling business around the discussion of the linear
%%   type of strings. Is this in section 3?}
where $\lceil \stringquote{} \rceil = I$ and
$\lceil c :: w \rceil = \literal c \otimes \lceil w \rceil$.

@@

The interpretation of dependent types as sets is standard
\cite{Hofmann_1997}. We present the concrete descriptions of the
semantics of linear types, as well as our non-standard non-linear
types in \Cref{fig:semantics}. The grammar for a literal $c$ has a
single parse precisely when the input string consists of the single
character. The grammar for the unit similarly has a single parse for
the empty string.  A parse of the tensor product $A \otimes B$
consists of a \emph{splitting} of the empty string into a prefix $w_1$
and suffix $w_2$ along with an $A$ parse of $w_1$ and $B$ parse of
$w_2$. A parse of $\bigoplus_{x:X} A$ is a pair of an element of the
set $X$ and a parse of $A(x)$, while dually a parse of $\bigwith_{x:X}
A$ is a \emph{function} taking any $x:X$ to a parse of $A(x)$. A
$w$-parse of $A \lto B$ is a function that takes an $A$ parse of some
other string $w'$ to a $B$ parse of $ww'$, and $B \tol A$ is the same
except the $B$ parse is for the reversed concatenation $w'w$.
%
The set $\uparrow A$ is the set of parse for the empty string for
$A$. This definition means that $\sem{\uparrow (A \lto B)}$ (or $\sem{\uparrow (B \tol A)}$) is
equivalent to the set of parse transformers:
\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
a parse in $\sem{A}$ that is mapped to the same parse by the parse transformers $\sem{f}$ and
$\sem{g}$.
%
The universe $L$ of linear types is interpreted as the set of all
small grammars.
\begin{figure}
  \begin{minipage}[t]{.5\textwidth}
  \begin{footnotesize}
    \begin{flalign*}
    &\sem{c}\gamma\,w = \{ c | w = c\}\\
    &\sem{I}\gamma\,w = \{ () \pipe w = \varepsilon \}\\
    &\sem{A \otimes B}\gamma\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    &\sem{A \lto B}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    &\sem{B \tol A}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}
  \end{flalign*}
  \end{footnotesize}
  \end{minipage}%
  \begin{minipage}[t]{.5\textwidth}
  \begin{footnotesize}
    \begin{flalign*}
    &\sem{\bigwith_{x:X} A}\gamma\,w = \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w \\
    &\sem{\uparrow A}\gamma = \sem{A}\gamma\,\varepsilon\\
    &\sem{\equalizer{a}{f}{g}}\gamma\,w = \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\

>>>>

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
%% \citestyle{acmauthoryear}   %% For author/year citations
\citestyle{acmnumeric}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{xspace}
\usepackage{thmtools} % needed for sane cleveref naming of theorems
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}

@@

\usepackage{listings}
\lstdefinelanguage{Agda}{
keywords={data, where, module, import, open, public,
record, field, let, in, if, then, else, case, of, with,
do, postulate, primitive, mutual, abstract, private,
forall, exists, cong, set, prop, sort, Level, Data, Type,
Renamer},
morekeywords=[2]{Set, Type, Prop, moduleFuncs, instance, Named},
sensitive=true,
comment=[l]{--},
morecomment=[s]{\{-}{-\}},
morestring=[b]",
mathescape=true,
escapeinside={(*@}{@*)}
}
\lstset{
language=Agda,
basicstyle=\ttfamily\footnotesize,
keywordstyle=\color{blue},
keywordstyle=[2]\color{teal},
identifierstyle=\color{black},
commentstyle=\color{gray}\textit,
stringstyle=\color{orange},
numbers=none,
numberstyle=\tiny\color{gray},
stepnumber=1,
numbersep=5pt,
showstringspaces=false,
tabsize=4,
captionpos=b,
breaklines=true,
breakatwhitespace=false,
rulecolor=\color{black},
texcl=true,
backgroundcolor=\color{backcolour},
frame=single
}

@@

\newcommand{\Subst}{\textrm{Subst}}
\newcommand{\SPF}{SPF}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\left\llbracket{#1}\right\rrbracket}
\newcommand{\semg}[1]{\sem{#1}\gamma}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}

\newcommand{\String}{\textbf{String}}
\newcommand{\Char}{\textbf{Char}}
\newcommand{\stringg}{\texttt{String}}
\newcommand{\charg}{\mathtt{Char}}
\newcommand{\StringSem}{\mathbf{String}}
\newcommand{\CharGram}{\texttt{Char}}
\newcommand{\StringGram}{\texttt{String}}
\newcommand{\stringNL}{String'}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\semcat}{\mathbf{C}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$\xspace}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

\begin{figure}
\begin{lstlisting}
f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
f (a , b) = inl (a , b)
\end{lstlisting}\begin{mathpar}
  \footnotesize
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {~}
      {a : \literal a \vdash a : \literal a}
      \\
      \inferrule
      {~}
      {b : \literal b \vdash b : \literal b}
    }
    {a : \literal a , b : \literal b \vdash (a , b) : \literal a \otimes
      \literal b}
  }
  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a , b) :
    (\literal a \otimes \literal b) \oplus \literal c}
\end{mathpar}
\caption{(\Agda{fillerlink}) \stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
}
\label{fig:fingram}
\end{figure}
For this interpretation of parse trees as terms to make sense, our
calculus cannot allow for \emph{any} of the usual structural rules of
type theory: weakening, contraction and exchange. Weakening allows
for variables to go unused, while contraction allows for the same
variable to be used twice, but in a parse tree, every character must
be accounted for exactly once. That is, we want to prevent the following
erroneous derivations,
\(
{a : \literal a , b : \literal b \not \vdash a : \literal a}\) and
\({a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a}
\).

Finally, the ordering of characters in a string cannot be ignored while
parsing, so we omit the exchange rule because it would allow
for variables in the context to be reordered. Likewise, we prevent the following derivation,
\(
  {a : \literal a , b : \literal b \not \vdash (b , a) : \literal b \otimes \literal a}
\).
% otherwise we would have that $\inl(x\otimes y)$ is a valid
% term in context $y:b,x:a$, but there should be no parses of $ba$ in
% this grammar.

@@

\begin{figure}
\begin{lstlisting}
data A(*@$^*$@*) : L where
  nil : (*@$\uparrow$@*)(A(*@$^*$@*))
  cons : (*@$\uparrow$@*)(A (*@$\lto$@*) A(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
\end{lstlisting}
\caption{Kleene Star as an inductive type}
\label{fig:kleenestarinductive}
\end{figure}

@@

\begin{figure}
\footnotesize
\begin{mathpar}
  \inferrule
  {
    \inferrule
    {~}
    {\cdot \vdash \nil : \A^*}
    \\
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {~}
        {a_1 : \A, a_2 : \A , as : \A^* \vdash \cons~a_1 (\cons~a_2~as) : \A^*}
      }
      {aa : \A \otimes \A , as : \A^* \vdash \letin {(a_1 , a_2)}
        {aa} {\cons~a_1 (\cons~a_2~as)} : \A^*}
    }
    {\cdot \vdash f := \lamblto {aa} {\lamblto {as} {\letin {(a_1 , a_2)}
        {aa} {\cons~a_1 (\cons~a_2~as)}}} : (\A \otimes \A) \lto \A^* \lto \A^*}
  }
  {aas : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(aas) : \A^* }
\end{mathpar}
\begin{lstlisting}
h : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
h nil = nil
h (cons (a1 , a2) as) = cons a1 (cons a2 (h as))
\end{lstlisting}
\caption{(\Agda{\fillerlink}) A parse transformer for abstract grammars}
\label{fig:kleeneabstractproof}
\end{figure}

@@

\begin{axiom}[Distributivity]
\label{ax:dist}
For any $A : (x : X) \to Y(x) \to L$, the definable distribuitivity function
%%%%%%%%%% An attempt to inline
% \(
%  {\LinSigTy{f}{\left(\PiTy{x}{X}{Y(x)}\right)}{\LinPiTy{x}{X}{A\,x\,(f\,x)}} \lto
% \LinPiTy{x}{X}{\LinSigTy{y}{Y(x)}{A\,x\,y}} \right)}
% \)
\(
 \lamblto{e}{\letin{\sigma\,f\,e'}{e}{\withlamb{x}{e'\,.\pi\,x}}} : \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
\LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}}
\)
has an inverse.
\end{axiom}

@@

  We demonstrate the weak equivalence between $Parse_{N}$ and
  $Parse_{D}$ by constructing parse transformers between the two
  grammars. To build the parse transformer
  $\ltonl{(Parse_{N} \lto Parse_{D})}$, we strengthen our inductive hypothesis to quantify over every start state
  and build a term
\newcommand{\NtoD}{\texttt{NtoD}}
\newcommand{\DtoN}{\texttt{DtoN}}
  \(
    \NtoD : \ltonl {\left(\Trace_{\N}~\s~\true \lto \LinPiTy{\X}{\D.\states}{\LinPiTy{\texttt{sInX}}{\X \ni \s}{ \Trace_{\D}~\X~\true}} \right) }
  \)
  that maps a trace in $\N$ from an arbitrary state $\s$ to a trace in $\D$
  that may begin at any subset of states $\X$ that contains $\s$. $\NtoD$ may then be instantiated at
  $\s = \N.\init$ and $\X = D.\mathsf{init}$ to get the desired
  parse transformer.

@@

To construct a term from DFA traces to NFA traces, we similarly strengthen our
inductive hypothesis and build a parse transformer that ranges over arbitrary $\epsilon$-closed subsets $X : \mathbb{P}_{\epsilon}{\N.\states}$,
\(
  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \LinSigTy{s}{\N.\states}{\LinSigTy{\texttt{sInX}}{\X \ni \s}{ \Trace_{\N}~s~\true}} \right)}
\).
Because the states of $D$ are $\epsilon$-closed subsets of $N.\states$, if two
states $s$ and $s'$ belong to the the same $\epsilon$-closed subset
$X : \mathbb{P}_{\epsilon}(N.\states)$ then there exists a $\epsilon$-path in
$N$ between the two, but we do not necessarily know which one. Similarly, the
data contained in the type $\Trace_{\D}~X~true$ is the \emph{existence} of an accepting trace in
$\N$ beginning at some state in $\D$.

To define $\DtoN$, we need a choice function that extracts out a trace in $N$
from the mere existence of one. We achieve this by choosing the smallest trace
through $\N$ subject to an ordering on the non-linear types $N.\states$,
$N.\transitions$, and $N.\epstransitions$. In essence, the given orderings
specify a global disambiguation strategy.
\end{proof}
%% For a moment, let us turn our attention to the content behind the traces in $\D$.
%% Traces from subset $\X$ to subset $\Y$ describes the \emph{existence} of some trace
%% in the $\N$ from some $\x \in \X$ to some $\y \in \Y$. In general it is not clear how
%% to extract out a choice of trace in $\N$, which is what $\DtoN$ is asking of us.
%% However, we may further take the states and transitions of $\N$ to be \emph{ordered}.
%% This allows us to
%% induce an order on traces in $\D$ from $\X$ to $\Y$. From these orderings we can then
%% define the necessary choice functions on each type by taking the smallest with
%% respect to the ordering.

@@

%% Next if the trace in $\D$ is defined via transition
%% $\X \overset{\stringquote{c}}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
%% $\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
%% $\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
%% there exists some $\s' \in \X$ such that $\s'$ transitions to $\s$ by some number of
%% $\epsilon$-transitions followed by a transition labeled by $\c$.
%% We order all
%% such traces of this form and choose the smallest one, which is then linearly
%% combined with the parse of $\c$ to build an $\N$ trace. This concludes the
%% definition of $\DtoN$, thus showing that $Parse_{N}$ and $Parse_{D}$ are
%% weakly equivalent.

Finally, given any regular expression we can construct a
\emph{strongly} equivalent NFA. While only weak equivalence is
required to construct a parser, proving the strong equivalence shows
that other aspects of formal grammar theory are also verifiable in
\theoryabbv.
\newcommand{\R}{\texttt{R}}
\begin{construction}[Thompson's Construction, \Agda{\fillerlink}]
  \label{cons:thompson}
  Given a regular expression $\R$, we build an NFA $\N$ such that $\R$ is
  strongly equivalent to $\Trace_{N}(\N.\init)$.
\end{construction}
\begin{proof}
We use a variant of Thompson's construction
\cite{thompsonProgrammingTechniquesRegular1968}, showing that NFAs are, up to strong equivalence, closed under each type operation for regular expressions.
\end{proof}
%% In this paper we will walk through
%% some of the cases of this proof. The remaining cases
%% follow
%% similar reasoning and the details may be found in the formalized artifact.

@@

%% In this sketch, we elide the proof that these maps do indeed form homomorphisms.
%% Further, we are eliding the needed NFAs for the cases where the regular
%% expression is $I$, $0$, a Kleene star, or a concatenation. All of these
%% elided details may be found in the Agda formalization.

@@

While we have shown only examples for context-free grammars, in fact
arbitrarily complex grammars are encodable in \theoryabbv. To
demonstrate this, we show that for any \emph{non-linear} function $P :
\StringSem \to U$, where here $\StringSem$ is the \emph{non-linear} type
of strings over the alphabet, we can construct a grammar whose parses
correspond to $P$.
%% Define $\StringSem$ to be the non-linear list type over the alphabet $\Sigma$ and
%% let  be a mapping from the set of strings to the universe of
%% nonlinear types. The function $P$ describes a family of nonlinear types
%% dependent on $\StringSem$.
%% We view the image of $P$ as a set of parse trees for each input string, and thus
%% $P$ carries the structure of a formal grammar. However, even though $P$ describes this grammatical data, it is not
%% a first-class linear type in \theoryabbv. 
\(
Reify~P = \LinSigTy{w} {\StringSem} {\LinSigTy {x} {P~w} {~\lceil w \rceil}}
\)
%% \steven{Probably put this ceiling business around the discussion of the linear
%%   type of strings. Is this in section 3?}
where $\lceil \stringquote{} \rceil = I$ and
$\lceil c :: w \rceil = \literal c \otimes \lceil w \rceil$.

@@

The interpretation of dependent types as sets is standard
\cite{Hofmann_1997}. We present the concrete descriptions of the
semantics of linear types, as well as our non-standard non-linear
types in \Cref{fig:semantics}. The grammar for a literal $c$ has a
single parse precisely when the input string consists of the single
character. The grammar for the unit similarly has a single parse for
the empty string.  A parse of the tensor product $A \otimes B$
consists of a \emph{splitting} of the empty string into a prefix $w_1$
and suffix $w_2$ along with an $A$ parse of $w_1$ and $B$ parse of
$w_2$. A parse of $\bigoplus_{x:X} A$ is a pair of an element of the
set $X$ and a parse of $A(x)$, while dually a parse of $\bigwith_{x:X}
A$ is a \emph{function} taking any $x:X$ to a parse of $A(x)$. A
$w$-parse of $A \lto B$ is a function that takes an $A$ parse of some
other string $w'$ to a $B$ parse of $ww'$, and $B \tol A$ is the same
except the $B$ parse is for the reversed concatenation $w'w$.
%
The set $\uparrow A$ is the set of parse for the empty string for
$A$. This definition means that $\sem{\uparrow (A \lto B)}$ (or $\sem{\uparrow (B \tol A)}$) is
equivalent to the set of parse transformers:
\( \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\).
